---
title: "Higher Education Outcomes"
author: "Ganesh kumar Chelladurai; Student ID 19233370"
date: "9/29/2019"
output: pdf_document
---

# Setup

```{r setup, echo=TRUE}
# this setup chunk loads the tidyverse. no other libraries should be used
library("tidyverse")

# setting echo=TRUE will cause all R code to be included in the pdf output
knitr::opts_chunk$set(echo=TRUE)
```

# Installing Latex

If you don't have Latex installed on your machine, you may see an
error when `knit`ting this file, and no pdf will be produced. Try
installing Latex as follows, and `knit` again.

```
# see https://yihui.name/tinytex/ for docs
> install.packages('tinytex')
> tinytex::install_tinytex()
```



# Loading data

First, load the data, and convert to a tibble (i.e. a `dplyr` dataframe) named `earnings`, with column names "Years.since.graduation", "NFQ.Level", "Sex", "Field", "Statistic", and "Value".

```{r}
#Importing tidyverse library
library(tidyverse)
#Using delimiter as semicolon and making table
#col_names adds column names to the data
#col_types defines the type of each column
earnings<- read_delim("earnings.csv",";",col_names=c("Years.since.graduation","NFQ.Level",
                                                     "Sex","Field","Statistic","Value"),
                                     col_types=cols(Years.since.graduation=col_integer(),
                                     NFQ.Level=col_factor(),Sex=col_factor(),
                                     Field=col_factor(),Statistic=col_factor(),
                                     Value=col_double()))

```

# Reshaping and cleaning

We should change the NFQ Level values to integers. The following function will be useful:

```{r, echo=TRUE}
convert_nfq <- function(s) {strtoi(substr(s, 11, 13))} # convert substring to int
```

Apply `convert_nfq` and check the result:
```{r}
# mutate_at changes the specific column specified "NFQ.Level" using function convert_nfq
# assigning the earnings back to earnings
earnings<-earnings%>%mutate_at(c("NFQ.Level"),convert_nfq)
```


Let's rename the `Years.since.graduation` column since it's a long name:

```{r}
# rename changes the name of the column rename(new_name = old_name)
# assigning the earnings back to earnings
earnings<-earnings%>%rename("Years"="Years.since.graduation")
```

Using `filter`, we discard all data where `Years` is not 1, because for some reason all that data is `NA`. Notice this reduces from 1600 rows to 400. 

```{r}
# assigning the earnings back to earnings
# Filter method filters the data based on the condition given(Years==1)
earnings<-earnings%>%filter(Years==1)
```

Our analysis is going to be based on Field, Sex, NFQ Level, Median Earnings, and Number of Graduates. We would like to have a column giving Median Earnings and another column giving Number of Graduates. That would be *tidy data*. Instead, we have one column giving the `Statistic` name, and another giving that statistic's `Value`. We fix this using `spread`. Notice that in the result, there
are several new columns. Some are shown directly, and the tibble says
"2 more variables" at the bottom.

```{r}
# spread splits the specified column (key,value) the key is split into the new columns 
# and the value of each column is the value specified
earnings<-earnings%>%spread(Statistic,Value)
```

Now we can discard the 25th and 75th percentiles and rename the other
columns:

```{r}
# select selects only the column specified .minus(-) avoids only the column specified 
# assigning the earnings back to earnings
earnings<-earnings%>%select(-c("P25 Earnings of Graduates (Euro)",
                               "P75 Earnings of Graduates (Euro)"))

#rename the columns
# assigning the earnings back to earnings
earnings<-earnings%>%rename("Median.Earnings"="P50 Earnings of Graduates (Euro)",
                            "Number.grads"="Number of Graduates (Persons)")
```

Now, let's have a summary of what we've got:
```{r}
summary(earnings)
```


# Plotting

Now we are ready to make a first plot. Let's look at the number of grads, by field and NFQ level.

```{r}
# plotting the graph with x as NFQ.level y as Field
# size varies the points plotted based on the column Number.grads
ggplot(earnings,mapping = aes(x=NFQ.Level,y=Field))+geom_point(aes(size=Number.grads))
```

Now we'll analyse the data by Sex. Getting the male and female dots to appear correctly is tricky, so here is a snippet you can add to your `ggplot` call:

`geom_point(position=position_nudge(x=0.25*(as.numeric(earnings$Sex) - 1.5)))`

```{r}
# plot graph with x as NFQ.Level and y as Field 
# colour assign different colur for each sex
# size varies the points plotted based on the column Number.grads
# position change the positon at which the point is plotted
ggplot(earnings,mapping = aes(x=NFQ.Level,y=Field,colour=Sex))+
  geom_point(aes(size=Number.grads),
             position=position_nudge(x=0.25*(as.numeric(earnings$Sex) - 1.5)))
```

Here is a more traditional scatter plot, but bear in mind that what we see are distributions of median earnings, not distributions of earnings. We will see a Warning message "Removed 17 rows containing missing values (geom_point)." - this is correct, of course, as we do have NA values for earnings wherever there were no grads. We can ignore it. 
```{r}
# plot graph with x as NFQ.Level and y as Median.Earnings 
# colour assign different colur for each sex
# size varies the points plotted based on the column Number.grads
# position change the positon at which the point is plotted
# guides is used to change the legends turning off the alplha legend with FALSE
ggplot(earnings,mapping = aes(x=NFQ.Level,y=Median.Earnings))+
  geom_point(aes(colour=Sex,size=Number.grads,alpha=0.5),
             position=position_nudge(x=0.25*(as.numeric(earnings$Sex) - 1.5)))+
             guides(alpha=FALSE)
```
Downloaded the  Earnings of Graduates by Sex, Field of Study, NFQ Level, Years since Graduation, Year and Statistic for the year 2011.
The same selection was made during the download expect I fileretered the no. of years after graduation to 1, saved the file as 'business_size.csv'.
URL: https://statbank.cso.ie/px/pxeirestat/Statire/SelectVarVal/Define.asp?maintable=HEO11&PLanguage=0

# Adding the column headers and filtering using year just like for earnings

```{r}
business<- read_csv("business_size.csv",
                    col_names=c("Years","Scale","Sex","Field","Number.grads"),
                    col_types=cols(Years=col_integer(),Scale=col_factor(),Sex=col_factor(),
                                   Field=col_factor(),Number.grads=col_double()))
business<-business%>%filter(Years==1)
```
# Format the existing dataset(earnings) to make fit for join
```{r}
# Removing unnecessary rows Number.grads,NFQ.Level
earnings<-earnings%>%select(-c(Number.grads,NFQ.Level,Sex,Years))
# group_by groups the tibble by both Field and sex 
# taking the mean of the Median.Earnings when we group the data 
# and made the column 2016(Earnings for 2016)
# na.rm remove the NA in the values
earnings2016<-earnings%>%group_by(Field)%>%summarise("2016"=mean(Median.Earnings, na.rm = TRUE))
```

#Our analysis will be based on the earnings between 2011 & 2016 based on Fields
#Reading and formatting the data
```{r}
#Reading the data, sama as above
earnings2011<- read_csv("earnings_2011.csv",
                        col_names=c("Years","NFQ.Level","Sex","Field","Earnings"),
                        col_types=cols(Years=col_integer(),NFQ.Level=col_factor(),
                                      Sex=col_factor(),Field=col_factor(),
                                      Earnings=col_double()))
#Removing unnecessary rows
earnings2011<-earnings2011%>%select(-c(Sex,NFQ.Level))
```
```{r}
# taking the mean of the Median.Earnings when we group the data
# and made the column 2011(Earnings for 2011)
# na.rm remove the NA in the values
earnings2011<-earnings2011%>%group_by(Field)%>%
  summarise("2011"=mean(Earnings, na.rm = TRUE))
```
```{r}
#Join the table based on field keep=FALSE removes the duplicate of column field 
#Donot need to keep field column from both tables
merged_table<-left_join(earnings2011,earnings2016,by=c("Field"="Field"),keep=FALSE)
```
```{r}
#Normalising the data using gather key=Year,Value=Salary,use columns 2011 and 2016
merged_table<-merged_table%>%gather(Year,Salary,"2011":"2016")
```
```{r}
#Plotting a bar graph with x as Field, y as Salary, fill apply colors to the year
#stat ="identity" represents the height of the bar to represent values
#lables is used to shortenen the labels as it is too long
#position_dodge is used to plot the bars separateley
ggplot(data=merged_table, aes(x=Field, y=Salary,fill=Year))+
  geom_bar(stat="identity",position=position_dodge())+scale_x_discrete(
      labels = c("Education"="Education","Arts and Humanities"="Arts",
      "Social Sciences, Journalism and Information"="Social\n&Journalism",
      "Business, Administration and Law"="Business\n&law",
      "Natural Sciences, Mathematics and Statistics"="Maths",
      "Information and Communication Technologies"="IT",
      "Engineering, Manufacturing and Construction"="Engineering",
      "Agriculture, Forestry, Fisheries and Veterinary"="Agro",
      "Health and Welfare"="Health"))
```

As we can see the mean earnings increased for every field for the year 2016 except for Education.
